Position Paper: Against Spurious Understanding---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}

### 2024-02-01 main.tex
fized typos

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Position Paper: Against Spurious Understanding: Dovelating Inflated AI Claims}

@@

\twocolumn[
\icmltitle{Position Paper: Against Spurious Understanding- 
           Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}

@@

Where AI used to be a computational modeling tool to better understand human cognition~\cite{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence. In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal. The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{bubeck2023sparks}. In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or showing performance considered `surprising' to humans.

For example, Google DeepMind claimed their AlphaGeometry model \cite{trinh2024geometry} reached a 'milestone' towards AGI. This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training. However, as noted by Hector Zenil \cite{hector2024linkedin}, work such as this had been initially introduced in the 1950s. Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system. Therefore, despite DeepMind's success in creating a powerfully fast geometry solving machine, it is still far from being AGI.

@@

When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness. AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\cite{values_in_ML}. As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder. In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.

Noticing these trends, we as the authors of this article are concerned. We feel that the current culture of racing towards Big Outcome Statements in industry and academic publishing too much disincentivizes efforts towards more thorough and nuanced actual problem understanding. At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims. With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs). In other words, it is both very tempting to claim surprising human-like achievements of AI, and as humans we are very prone to genuinely believe this.

To strengthen our argument, in the remainder of this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made. We work out different experiments which may invite similar claims on models yielding more intelligent outcomes than would have been expected, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate how we feel these claims should \emph{not} be made. Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there. Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.

@@

For each layer, we compute linear probes through Ridge regression on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity. To allow for a comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels. To measure price inflatio,n we use percentage changes proxied by log differences.

@@

%\subsubsection{Stochastic Parrots AfterAll?}\label{stochastic-parrots-after-all}
\subsubsection{Sparks of economic understanding?}\label{stochastic-parrots-after-all}
%The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive:
Even though \emph{FOMC-RoBERTa} (which is substantially smaller than the models tested in~\citet{gurnee2023languagev2} was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions). So, have we uncovered further evidence that LLMs ``aren't mere stochastic
parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
of the economy just by `reading' central bank communications? Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?

@@

\section{On Human Proneness to Over-Interpretation}
Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}. Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}. Our examples demonstrate, that this is achievable even for small models, while these models have certainly not developed an intrinsic ``understanding'' of the world. Thus, to draw conclusions about the emerging capabilities of AI models, we argue that more conservative and rigorous tests are needed.

@@

Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects. These might include the weather and other natural forces, pets and other animals, and gadgets and other pieces of technology \cite{epley2007seeing}. Formally studied as early as 1944, \cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative. 
%Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review). 
Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems. As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.

@@

These pressures extend to expert AI practitioners and researchers as well as laypersons. In both scholarly and commercial fields, AI experts face considerable pressures to demonstrate competence in AI-related work. Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g. \cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall scholarly publications \footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }. This is also evident in peer-reviewed publications \cite{Maslej2023-pi}, with both more than doubling from 2010 to 2020. The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}. Outside of academia, the number of jobs requiring AI expertise have increased several fold, with demand for 'Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}. Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.

@@

As a consequence, we call for the community to create explicit room for organized skepticism. For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed. If claims of human-like or superhuman intelligence are made, these should be subject to severe tests that go beyond display of surprise. Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activity as a community may help in incentivizing and normalizing constructive adversarial questioning.

In this, as the quest for AGI so deeply roots in human-like recognition, we also want to put an explicit word of warning about use of terminology. Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have common understanding in specialized research communities, but have bigger, anthropormophic connotations in laypersons. We add our voice to emerging calls for this~\cite{shanahan2024talking}.

@@

Thus, we as academic researchers carry great responsibility on how the narrative will unfold, and what claims are believed. We call upon our colleagues to be explicitly mindful on this. As attractive it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager in running with made claims.

>>>>

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Position Paper: Against Spurious Understanding---Dovelating Inflated AI Claims}

@@

\twocolumn[
\icmltitle{Position Paper: Against Spurious Understanding---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}

@@

Where AI used to be a computational modeling tool to better understand human cognition~\cite{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence. In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal. The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{bubeck2023sparks}. In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.

For example, Google DeepMind claimed their AlphaGeometry model~\cite{trinh2024geometry} reached a `milestone' towards AGI. This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training. However, as noted by Hector Zenil~\cite{hector2024linkedin}, work such as this had been initially introduced in the 1950s. Without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system. Therefore, despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.

@@

When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that aims towards human-like problem-solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience\footnote{\url{https://www.scientificamerican.com/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/}} and consciousness. AI technology only becomes more difficult to understand as academic publishing in the space largely favors performance, generalization, quantitative evidence, efficiency, building on past work, and novelty~\cite{values_in_ML}. As such, works that make it into top-tier venues tend to propose heavier and more complicated technical takes on tasks that (in the push towards generalizability) get more vague, while the scaling-up of data makes traceability of possible memorization harder. In a submission-overloaded reality, researchers may further get incentivized to oversell and overstate achievement claims.

Noticing these trends, we as the authors of this article are concerned. We feel that the current culture of racing toward Big Outcome Statements in industry and academic publishing too much disincentivizes efforts toward more thorough and nuanced actual problem understanding. At the same time, as the outside world is so eager to adopt AI technology, (too) strong claims make for good sales pitches, but a question is whether there is indeed sufficient evidence for these claims. With successful A(G)I outcomes needing to look human-like, this also directly plays into risks of anthropomorphizing (the attribution of human-like qualities to non-human objects) and confirmation bias (the seeking-out and/or biased interpretation of evidence in support of one's beliefs). In other words, it is very tempting to claim surprising human-like achievements of AI, and as humans, we are very prone to genuinely believe this.

To strengthen our argument, in the remainder of this paper, we first consider a recently viral work~\cite{gurnee2023languagev2} in which claims about the learning of world models by LLMs were made. We work out different experiments that may invite similar claims on models yielding more intelligent outcomes than would have been expected, especially of a nature that they may look attractive for adoption as an innovative predictive tool in concrete application domains---while we at the same time indicate how we feel these claims should \emph{not} be made. Furthermore, we present a review of social science findings that underline how prone humans are to being enticed by patterns that are not really there. Combining this with the way in which media portrayal of AI has tended towards science-fiction imagery of mankind-threatening robots, we argue that the current AI culture is a perfect storm for making and believing inflated claims, and call upon our fellow academics to be extra mindful and scrutinous about this.

@@

For each layer, we compute linear probes through Ridge regression on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity. To allow for a comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels. To measure price inflation we use percentage changes proxied by log differences.

@@

%\subsubsection{Stochastic Parrots AfterAll?}\label{stochastic-parrots-after-all}
\subsubsection{Sparks of economic understanding?}\label{stochastic-parrots-after-all}
%The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive:
Even though \emph{FOMC-RoBERTa}, which is substantially smaller than the models tested in~\citet{gurnee2023languagev2}, was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions). So, have we uncovered further evidence that LLMs ``aren't mere stochastic
parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
of the economy just by `reading' central bank communications? Thus, can economists readily adopt FOMC-RoBERTa as a domain-relevant tool?

@@

\section{On Human Proneness to Over-Interpretation}

Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}. Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}. Our examples demonstrate that this is achievable even for small models, while these models have certainly not developed an intrinsic ``understanding'' of the world. Thus, to draw conclusions about the emerging capabilities of AI models, we argue that more conservative and rigorous tests are needed.

@@

Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects. These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\cite{epley2007seeing}. Formally studied as early as 1944,~\cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative. 
%Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review). 
Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems. As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.

@@

These pressures extend to expert AI practitioners and researchers as well as laypersons. In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in AI-related work. Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g. \cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall scholarly publications \footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }. This is also evident in peer-reviewed publications \cite{Maslej2023-pi}, with both more than doubling from 2010 to 2020. The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}. Outside of academia, the number of jobs requiring AI expertise has increased severalfold, with demand for 'Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}. Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.

@@

As a consequence, we call for the community to create explicit room for organized skepticism. For research that seeks to explain a phenomenon, clear hypothesis articulation and strong null hypothesis formulation will be needed. If claims of human-like or superhuman intelligence are made, these should be subject to severe tests that go beyond the display of surprise. Apart from focusing on getting novel improvements upon state-of-the-art published, organizing red-teaming activities as a community may help in incentivizing and normalizing constructive adversarial questioning.

In this, as the quest for AGI is so deeply rooted in human-like recognition, we also want to put an explicit word of warning about the use of terminology. Many terms used in current A(G)I research (e.g.\ emergence, intelligence, learning, `better than human' performance) have a common understanding in specialized research communities, but have bigger, anthropomorphic connotations in laypersons. We add our voice to emerging calls for this~\cite{shanahan2024talking}.

@@

Thus, we as academic researchers carry great responsibility for how the narrative will unfold, and what claims are believed. We call upon our colleagues to be explicitly mindful of this. As attractive as it may be to beat the state-of-the-art with a grander claim, let us return to the Mertonian norms, and thus safeguard our academic legitimacy in a world that only will be eager to run with made claims.
